---
title: "Project_final"
author: "Mahalakshmi Aranganathan"
date: "2025-04-17"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
chooseCRANmirror(graphics = FALSE, ind = 1)
tinytex::install_tinytex(force = TRUE)
```

```{r}
# 📦 Install required package (only needs to be run once)
# install.packages("readr")  # Uncomment this line if not already installed

# 📚 Load libraries
library(readr)    # For reading CSV files
library(dplyr)    # For data manipulation

# 📁 Read the clinical data CSV file
clinical_combined_data <- read_csv("clinical_combined_data.csv")

# 👀 View the dataset
View(clinical_combined_data)

# 🧹 Remove duplicate columns
clinical_combined_data <- clinical_combined_data[, !duplicated(colnames(clinical_combined_data))]

# 🧹 Remove columns with suffixes ".x" or ".y" (from merged datasets)
clinical_combined_data <- clinical_combined_data[, !grepl("\\.x$|\\.y$", colnames(clinical_combined_data))]

# 🧹 Remove columns that contain only "--" as values
clinical_combined_data <- clinical_combined_data[, !apply(clinical_combined_data, 2, function(x) all(x == "--"))]

# 🏷️ Display remaining column names
colnames(clinical_combined_data)

# 🔢 Convert 'age_at_diagnosis' from days to years and round to 2 decimals
clinical_combined_data <- clinical_combined_data %>%
  mutate(age_at_diagnosis = as.numeric(as.character(age_at_diagnosis)) / 365.25) %>%
  mutate(age_at_diagnosis = round(age_at_diagnosis, 2))

# 👀 View the updated dataset
View(clinical_combined_data)

# 🔄 Replace 'age_at_diagnosis' with 'age_at_index' (renaming column)
clinical_combined_data <- clinical_combined_data %>%
  select(-age_at_diagnosis) %>%
  rename(age_at_diagnosis = age_at_index)

# 🧬 Convert selected variables to categorical (factors)
clinical_combined_data <- clinical_combined_data %>% 
  mutate(
    ajcc_pathologic_stage = ifelse(ajcc_pathologic_stage %in% c("--", "'--"), "unknown", ajcc_pathologic_stage),
    exposure_type = as.factor(exposure_type),
    tobacco_smoking_status = as.factor(tobacco_smoking_status),
    gender = as.factor(gender),
    race = as.factor(race),
    vital_status = as.factor(vital_status),
    treatment_outcome = as.factor(treatment_outcome),
    ajcc_pathologic_stage = factor(
      ajcc_pathologic_stage,
      levels = c("Stage IIB", "Stage IB", "Stage IIIA", "Stage IIA", "Unknown", "Stage IA", "Stage IIIB", "Stage IV", "Stage I", "Stage                      II", "Stage III"),
      labels = c("Stage IIB", "Stage IB", "Stage IIIA", "Stage IIA", "Unknown", "Stage IA", "Stage IIIB", "Stage IV", "Stage I", "Stage II", "Stage III")
    )
  )


# ❌ Replace invalid entries ("'--") with NA
clinical_combined_data[clinical_combined_data == "'--"] <- NA

# 👀 View cleaned data
View(clinical_combined_data)

# 📊 Check missing values per column
colSums(is.na(clinical_combined_data))

# ⚠️ Set threshold: Remove columns with more than 70% missing values
threshold <- 0.7
missing_percent <- colSums(is.na(clinical_combined_data)) / nrow(clinical_combined_data)

# 🧹 Keep only columns with <70% missing data
clinical_combined_data <- clinical_combined_data %>%
  select(which(missing_percent < threshold))

# 👀 View after filtering
View(clinical_combined_data)

# 🔄 Convert 'pack_years_smoked' to numeric
clinical_combined_data <- clinical_combined_data %>%
  mutate(pack_years_smoked = as.numeric(as.character(pack_years_smoked)))

# 🧪 Impute missing numeric values with median
clinical_combined_data <- clinical_combined_data %>%
  mutate(
    pack_years_smoked = ifelse(is.na(pack_years_smoked), median(pack_years_smoked, na.rm = TRUE), pack_years_smoked),
    age_at_diagnosis = ifelse(is.na(age_at_diagnosis), median(age_at_diagnosis, na.rm = TRUE), age_at_diagnosis)
  )

# 📈 Define function to compute mode (most frequent value)
mode_func <- function(x) {
  uniq_vals <- unique(na.omit(x))
  uniq_vals[which.max(tabulate(match(x, uniq_vals)))]
}

# 🧩 Impute missing categorical values using mode
clinical_combined_data <- clinical_combined_data %>%
  mutate(
    tobacco_smoking_status = ifelse(is.na(tobacco_smoking_status), mode_func(tobacco_smoking_status), tobacco_smoking_status),
    treatment_type = ifelse(is.na(treatment_type), mode_func(treatment_type), treatment_type)
  )

# 👀 Final cleaned dataset
View(clinical_combined_data)

# 🏷️ Display column names of final dataset
colnames(clinical_combined_data)

```

```{r}
library(dplyr)
library(caret) # For createDataPartition function
library(randomForest) # For random forest model

# Function to get mode of a vector (for imputing missing values)
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Select relevant columns
selected_data <- clinical_combined_data %>%
  select(vital_status, ajcc_pathologic_stage, treatment_type, gender, race)

# Impute missing values (for categorical variables like treatment_type, gender, race, etc.)
selected_data$treatment_type[is.na(selected_data$treatment_type)] <- as.character(getmode(selected_data$treatment_type))
selected_data$gender[is.na(selected_data$gender)] <- as.character(getmode(selected_data$gender))
selected_data$race[is.na(selected_data$race)] <- as.character(getmode(selected_data$race))
selected_data$ajcc_pathologic_stage[is.na(selected_data$ajcc_pathologic_stage)] <- as.character(getmode(selected_data$ajcc_pathologic_stage))

# Remove any rows with missing vital_status
selected_data <- selected_data %>%
  na.omit()

# Check the number of levels for each factor variable
sapply(selected_data, function(x) if(is.factor(x)) length(levels(x)) else NA)

# Split Data into training and testing sets
set.seed(42)
train_index <- createDataPartition(selected_data$vital_status, p = 0.8, list = FALSE)
train_data <- selected_data[train_index, ]
test_data <- selected_data[-train_index, ]

# Ensure test_data has the same factor levels as train_data for all categorical variables
for(col in names(selected_data)) {
  if(is.factor(selected_data[[col]])) {
    levels(test_data[[col]]) <- levels(train_data[[col]])
  }
}

# Ensure vital_status is a factor variable
train_data$vital_status <- factor(train_data$vital_status)
test_data$vital_status <- factor(test_data$vital_status)

# Fit the Logistic regression Model
log_model <- glm(vital_status ~ ., data = train_data, family = binomial)

# Evaluate the Logistic regression Model
predictions_log <- predict(log_model, test_data, type = "response")
predicted_classes_log <- ifelse(predictions_log > 0.5, levels(train_data$vital_status)[1], levels(train_data$vital_status)[2]) 

# Confusion Matrix for Logistic regression
conf_matrix_log <- table(predicted_classes_log, test_data$vital_status)
print(conf_matrix_log)

# Print predicted classes and actual values for Logistic regression
print(predicted_classes_log)
print(test_data$vital_status)

# Train a Random Forest Model
rf_model <- randomForest(vital_status ~ ., data = train_data)

# Evaluate the Random Forest Model
predicted_classes_rf <- predict(rf_model, newdata = test_data)

# Confusion Matrix for Random Forest
conf_matrix_rf <- table(predicted_classes_rf, test_data$vital_status)
print(conf_matrix_rf)

# Print predicted classes and actual values for Random Forest
print(predicted_classes_rf)
print(test_data$vital_status)

confusion_log <- confusionMatrix(conf_matrix_log)
print(confusion_log)

confusion_rf <- confusionMatrix(conf_matrix_rf)
print(confusion_rf)


```

```{r}
# Load necessary libraries
library(dplyr)
library(caret) # For createDataPartition function
library(rpart) # For Decision Tree model
library(randomForest) # For Random Forest model
library(pROC) # For ROC curve and AUC
library(e1071) # For calculating performance metrics

# Select relevant columns and handle missing values
selected_data <- clinical_combined_data %>%
  select(vital_status, ajcc_pathologic_stage, treatment_type, gender, race) %>%
  na.omit()
View(selected_data)
levels(selected_data$vital_status)
# Split Data into training and testing sets
set.seed(42)
train_index <- createDataPartition(selected_data$vital_status, p = 0.8, list = FALSE)
train_data <- selected_data[train_index, ]
test_data <- selected_data[-train_index, ]

# Fit the Logistic Regression Model
log_model <- glm(vital_status ~ ., data = train_data, family = binomial)

# Predict using Logistic Regression
predictions_log <- predict(log_model, test_data, type = "response")
predicted_classes_log <- ifelse(predictions_log > 0.5, "Alive", "Dead")

# Fit the Random Forest Model
rf_model <- randomForest(vital_status ~ ., data = train_data)

# Predict using Random Forest
predicted_classes_rf <- predict(rf_model, newdata = test_data)

# Fit the Decision Tree Model
tree_model <- rpart(vital_status ~ ., data = train_data, method = "class")

# Predict using Decision Tree
predicted_classes_tree <- predict(tree_model, test_data, type = "class")

evaluate_model <- function(predicted_classes, actual_classes, model_name) {
  # Ensure predicted_classes and actual_classes are factors with the same levels
  predicted_classes <- factor(predicted_classes, levels = c("Dead", "Alive"))
  actual_classes <- factor(actual_classes, levels = c("Dead", "Alive"))
  
  # Confusion matrix
  conf_matrix <- table(predicted_classes, actual_classes)
  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
  sensitivity <- conf_matrix[2, 2] / sum(conf_matrix[2, ])
  specificity <- conf_matrix[1, 1] / sum(conf_matrix[1, ])
  precision <- conf_matrix[2, 2] / sum(conf_matrix[, 2])
  f1_score <- 2 * (precision * sensitivity) / (precision + sensitivity)
  
  # AUC: Convert factor to numeric (Dead = 0, Alive = 1)
  auc <- roc(as.numeric(actual_classes) - 1, as.numeric(predicted_classes) - 1)$auc
  
  # Print metrics
  cat("\nModel: ", model_name, "\n")
  cat("Confusion Matrix:\n")
  print(conf_matrix)
  cat("Accuracy: ", accuracy, "\n")
  cat("Sensitivity (Recall): ", sensitivity, "\n")
  cat("Specificity: ", specificity, "\n")
  cat("Precision: ", precision, "\n")
  cat("F1-Score: ", f1_score, "\n")
  cat("AUC: ", auc, "\n")
}

# Evaluate Logistic Regression
evaluate_model(predicted_classes_log, test_data$vital_status, "Logistic Regression")

# Evaluate Random Forest
evaluate_model(predicted_classes_rf, test_data$vital_status, "Random Forest")

# Evaluate Decision Tree
evaluate_model(predicted_classes_tree, test_data$vital_status, "Decision Tree")

# ###The evaluation results for the three models are as follows:
# 
# 1. Logistic Regression:
# Accuracy: 0.31 (Low)
# 
# Sensitivity (Recall): 0.30 (Low)
# 
# Specificity: 0.33 (Low)
# 
# Precision: 0.43 (Moderate)
# 
# F1-Score: 0.35 (Low)
# 
# AUC: 0.67 (Moderate)
# 
# 2. Random Forest:
# Accuracy: 0.77 (Good)
# 
# Sensitivity (Recall): 0.75 (High)
# 
# Specificity: 0.79 (High)
# 
# Precision: 0.72 (High)
# 
# F1-Score: 0.74 (Good)
# 
# AUC: 0.77 (Good)
# 
# 3. Decision Tree:
# Accuracy: 0.71 (Good)
# 
# Sensitivity (Recall): 0.69 (Moderate)
# 
# Specificity: 0.72 (Moderate)
# 
# Precision: 0.61 (Moderate)
# 
# F1-Score: 0.65 (Moderate)
# 
# AUC: 0.70 (Moderate)
# 
# Key Observations:
# Random Forest consistently outperforms the other two models across all evaluation metrics. It has the highest accuracy, sensitivity, specificity, precision, F1-score, and AUC.
# 
# Logistic Regression shows the poorest performance, with low accuracy and sensitivity. Its precision is also lower compared to the others, indicating that it struggles with classification.
# 
# Decision Tree performs decently but does not match Random Forest in terms of accuracy, sensitivity, or precision.
# 
# Conclusion:
# Random Forest is the most reliable model in this case, achieving the best results across all metrics.
# 
# Decision Tree performs better than Logistic Regression but is not as strong as Random Forest.
# 
# Logistic Regression shows limited performance and may not be suitable for this dataset based on the current evaluation.
# ###
```

```{r}
#Bar Plot for Model Comparison

performance_metrics <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "Decision Tree"),
  Accuracy = c(0.31, 0.77, 0.71),
  Sensitivity = c(0.30, 0.75, 0.69),
  Specificity = c(0.33, 0.79, 0.72),
  Precision = c(0.43, 0.72, 0.61),
  F1_Score = c(0.35, 0.74, 0.65),
  AUC = c(0.67, 0.77, 0.70)
)

#Reshape the data frame to long format for ggplot
library(tidyr)
performance_metric_long <- performance_metrics %>%
  gather(key = "Metric", value = "Value", -Model)

#Create a bar plot using ggplot2
library(ggplot2)
ggplot(performance_metric_long, aes(x = Model, y = Value, fill = Metric)) + geom_bar(stat = "identity", position = "dodge") + 
  theme_minimal() + labs (title = "Model Performance Comparision", y = "Score", x = "Model") + theme(axis.test.x = element_text(angle = 45, hjust = 1)) + 
scale_fill_brewer(palette = "Set3")

#ROC Curve for AUC Comparison

library(pROC)

#Generate ROC curves for each model
roc_log <- roc(test_data$vital_status, as.numeric(predictions_log))
roc_rf <- roc(test_data$vital_status, as.numeric(predicted_classes_rf))
roc_tree <- roc(test_data$vital_status, as.numeric(predicted_classes_tree))

#Plot ROC curves

plot(roc_log, col = "red", main = "ROC Curves Comparision", lwd = 2)
lines(roc_rf, col = "blue", lwd = 2)
lines(roc_tree, col = "green", lwd =2)

#Add Legend
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "Decision Tree"), col = c("red","blue","green"), lwd = 2)

#HEATMAP FOR CORRELATION BETWEEN MODEL METRICS
# Create a correlation matrix for the models
cor_matrix <- cor(performance_metrics[, -1])  # Exclude 'Model' column

# Plot heatmap
install.packages("pheatmap")
library(pheatmap)
pheatmap(cor_matrix, cluster_rows = TRUE, cluster_cols = TRUE, display_numbers = TRUE)


#BOXPLOT
# Data frame to hold evaluation metrics for boxplot
metrics_data <- data.frame(
  Model = rep(c("Logistic Regression", "Random Forest", "Decision Tree"), each = 6),
  Accuracy = c(0.31, 0.77, 0.71),
  Sensitivity = c(0.30, 0.75, 0.69),
  Specificity = c(0.33, 0.79, 0.72),
  Precision = c(0.43, 0.72, 0.61),
  F1_Score = c(0.35, 0.74, 0.65),
  AUC = c(0.67, 0.77, 0.70)
)

# Reshape data for ggplot
metrics_data_long <- metrics_data %>%
  gather(key = "Metric", value = "Value", -Model)

# Create boxplot
ggplot(metrics_data_long, aes(x = Model, y = Value, fill = Model)) +
  geom_boxplot() +
  facet_wrap(~ Metric, scales = "free") +
  theme_minimal() +
  labs(title = "Performance Comparison of Models")


```

```{r}
# Get the feature importance from the Random Forest model
importance_rf <- randomForest::importance(rf_model)

# Display the importance
print(importance_rf)

# If you want to display the importance in a more readable format
importance_rf_df <- data.frame(Feature = rownames(importance_rf), Importance = importance_rf[,1])
importance_rf_df <- importance_rf_df[order(-importance_rf_df$Importance), ]  # Sort by importance

# View the sorted importance
print(importance_rf_df)

# Visualize the feature importance using ggplot2
library(ggplot2)

# Create a bar plot
ggplot(importance_rf_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Feature Importance in Random Forest Model",
       x = "Features",
       y = "Importance") +
  theme_minimal()

```

```{r}
# Load necessary library
library(dplyr)

# View structure of vital_status column
str(clinical_combined_data$vital_status)

# Convert vital_status to factor (if not already)
clinical_combined_data$vital_status <- as.factor(clinical_combined_data$vital_status)

# Convert age_at_diagnosis to numeric
clinical_combined_data$age_at_diagnosis <- as.numeric(clinical_combined_data$age_at_diagnosis)

clinical_combined_data$pack_years_smoked <- as.numeric(as.character(clinical_combined_data$pack_years_smoked))


# Recode tobacco_smoking_status into meaningful categories
clinical_combined_data <- clinical_combined_data %>%
  mutate(tobacco_smoking_status = recode(as.numeric(tobacco_smoking_status),
                                         `1` = "Reformed Smoker",
                                         `2` = "Reformed Smoker",
                                         `3` = "Non-Smoker",
                                         `4` = "Current Smoker",
                                         `5` = "Unknown",
                                         `6` = "Reformed Smoker",
                                         `7` = "Unknown"),
         tobacco_smoking_status = factor(tobacco_smoking_status))

# Convert pack_years_smoked to factor (assuming it's categorical)
clinical_combined_data$pack_years_smoked <- as.factor(clinical_combined_data$pack_years_smoked)

# Remove rows with missing pack_years_smoked
clinical_combined_data <- clinical_combined_data[!is.na(clinical_combined_data$pack_years_smoked), ]

# View summaries and tables
table(clinical_combined_data$tobacco_smoking_status)
table(clinical_combined_data$exposure_type)
summary(clinical_combined_data$pack_years_smoked)
summary(clinical_combined_data$age_at_diagnosis)

# Check if vital_status is binary before fitting logistic regression
table(clinical_combined_data$vital_status)

# Fit logistic regression model
model <- glm(vital_status ~ pack_years_smoked + tobacco_smoking_status, 
             family = binomial, data = clinical_combined_data)
summary(model)

# Check factor levels
levels(clinical_combined_data$pack_years_smoked)
levels(clinical_combined_data$tobacco_smoking_status)
levels(clinical_combined_data$vital_status)

# View cleaned dataset (interactive RStudio only)
# View(clinical_combined_data)

```

```{r}
# Load necessary libraries
library(dplyr)
library(caret)         # For createDataPartition
library(rpart)         # For Decision Tree
library(randomForest)  # For Random Forest
library(pROC)          # For ROC curve and AUC
library(e1071)         # For performance metrics

# Select relevant columns and handle missing values
selected_data <- clinical_combined_data %>%
  select(vital_status, ajcc_pathologic_stage, treatment_type, gender, race, tobacco_smoking_status) %>%
  na.omit()

# Recode vital_status to numeric: 0 = Alive, 1 = Dead
selected_data$vital_status <- ifelse(selected_data$vital_status == "Alive", 0, 1)
selected_data$vital_status <- as.factor(selected_data$vital_status)

# Split into training and testing sets
set.seed(42)
train_index <- createDataPartition(selected_data$vital_status, p = 0.8, list = FALSE)
train_data <- selected_data[train_index, ]
test_data <- selected_data[-train_index, ]

# ---------------- Logistic Regression ----------------
log_model <- glm(vital_status ~ ., data = train_data, family = binomial)
predictions_log <- predict(log_model, newdata = test_data, type = "response")
predicted_classes_log <- ifelse(predictions_log > 0.5, 1, 0)
predicted_classes_log <- as.factor(predicted_classes_log)

# ---------------- Random Forest ----------------
rf_model <- randomForest(vital_status ~ ., data = train_data)
predicted_classes_rf <- predict(rf_model, newdata = test_data)
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[, "1"]

# ---------------- Decision Tree ----------------
tree_model <- rpart(vital_status ~ ., data = train_data, method = "class")
predicted_classes_tree <- predict(tree_model, newdata = test_data, type = "class")

# ---------------- Evaluation Function ----------------
evaluate_model <- function(predicted_classes, actual_classes, predicted_probs = NULL, model_name = "") {
  predicted_classes <- factor(predicted_classes, levels = c(0, 1))
  actual_classes <- factor(actual_classes, levels = c(0, 1))
  
  conf_matrix <- table(Predicted = predicted_classes, Actual = actual_classes)
  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
  sensitivity <- conf_matrix["1", "1"] / sum(conf_matrix[, "1"])
  specificity <- conf_matrix["0", "0"] / sum(conf_matrix[, "0"])
  precision <- conf_matrix["1", "1"] / sum(conf_matrix["1", ])
  f1_score <- 2 * (precision * sensitivity) / (precision + sensitivity)
  
  cat("\nModel: ", model_name, "\n")
  cat("Confusion Matrix:\n")
  print(conf_matrix)
  cat("Accuracy: ", round(accuracy, 4), "\n")
  cat("Sensitivity (Recall): ", round(sensitivity, 4), "\n")
  cat("Specificity: ", round(specificity, 4), "\n")
  cat("Precision: ", round(precision, 4), "\n")
  cat("F1-Score: ", round(f1_score, 4), "\n")
  
  if (!is.null(predicted_probs)) {
    roc_obj <- roc(actual_classes, predicted_probs)
    cat("AUC: ", round(auc(roc_obj), 4), "\n")
    plot(roc_obj, col = "blue", main = paste("ROC Curve -", model_name))
  }
}

# ---------------- Evaluate Models ----------------
evaluate_model(predicted_classes_log, test_data$vital_status, predictions_log, "Logistic Regression")
evaluate_model(predicted_classes_rf, test_data$vital_status, rf_probs, "Random Forest")
evaluate_model(predicted_classes_tree, test_data$vital_status, NULL, "Decision Tree")


```

```{r}
# Get the feature importance from the Random Forest model
importance_rf <- randomForest::importance(rf_model)

# Display the importance
print(importance_rf)

# If you want to display the importance in a more readable format
importance_rf_df <- data.frame(Feature = rownames(importance_rf), Importance = importance_rf[,1])
importance_rf_df <- importance_rf_df[order(-importance_rf_df$Importance), ]  # Sort by importance

# View the sorted importance
print(importance_rf_df)

# Visualize the feature importance using ggplot2
library(ggplot2)

# Create a bar plot
ggplot(importance_rf_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Feature Importance in Random Forest Model",
       x = "Features",
       y = "Importance") +
  theme_minimal()

```

#To understand which treatment model led to a the positive effect more
```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(broom)

# Logistic regression with only treatment_type
log_model_treatment <- glm(vital_status ~ treatment_type, data = selected_data, family = binomial)

# Tidy the model output
tidy_model <- tidy(log_model_treatment) %>%
  filter(term != "(Intercept)") %>%
  mutate(treatment = gsub("treatment_type", "", term))

# Create bar plot of treatment effects
ggplot(tidy_model, aes(x = reorder(treatment, estimate), y = estimate, fill = estimate > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  labs(title = "Effect of Treatment Type on Survival",
       x = "Treatment Type",
       y = "Estimate (Log-Odds)") +
  scale_fill_manual(values = c("red", "green"), labels = c("Negative", "Positive"), name = "Effect") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))

```

```{r}
# Install and load required packages if not already installed
required_packages <- c("survival", "survminer", "dplyr", "ggplot2")

# Check for missing packages and install them
installed_packages <- installed.packages()[, "Package"]
to_install <- setdiff(required_packages, installed_packages)

if(length(to_install) > 0) {
  install.packages(to_install)
}

# Load libraries
library(survival)
library(survminer)
library(dplyr)
library(ggplot2)

# Select relevant columns and preprocess data for survival analysis
surv_data <- clinical_combined_data %>%
  select(
    days_to_death, 
    days_to_diagnosis, 
    vital_status, 
    age_at_diagnosis, 
    ajcc_pathologic_stage, 
    tobacco_smoking_status, 
    pack_years_smoked, 
    gender
  ) %>%
  mutate(
    # Ensure days_to_death and days_to_diagnosis are numeric
    days_to_death = as.numeric(days_to_death),
    days_to_diagnosis = as.numeric(days_to_diagnosis)
  )

# Function to calculate the mode (most frequent value)
get_mode <- function(x) {
  ux <- unique(na.omit(x))  # Remove NAs and find unique values
  ux[which.max(tabulate(match(x, ux)))]  # Return the most frequent value
}

# Impute missing values for 'days_to_death' and 'days_to_diagnosis'
surv_data$days_to_death[is.na(surv_data$days_to_death)] <- get_mode(surv_data$days_to_death)
surv_data$days_to_diagnosis[is.na(surv_data$days_to_diagnosis)] <- get_mode(surv_data$days_to_diagnosis)

# Define event status (1 = Dead, 0 = Alive) and calculate survival time
surv_data <- surv_data %>%
  mutate(
    event = ifelse(vital_status == "Dead", 1, 0),
    survival_time = pmax(days_to_death - days_to_diagnosis, 1)  # Ensure survival time is positive
  )

# Remove rows with missing or undefined AJCC stage
surv_data <- surv_data %>%
  filter(!is.na(ajcc_pathologic_stage) & ajcc_pathologic_stage != "") %>%
  mutate(ajcc_pathologic_stage = factor(ajcc_pathologic_stage))

# Create survival object
surv_obj <- Surv(time = surv_data$survival_time, event = surv_data$event)

# Fit Kaplan-Meier survival curves based on AJCC pathological stage
km_fit <- survfit(surv_obj ~ ajcc_pathologic_stage, data = surv_data)

# Determine number of levels in AJCC Pathologic Stage
num_levels <- length(levels(surv_data$ajcc_pathologic_stage))

# Choose an appropriate color palette based on the number of levels
palette_choice <- if(num_levels <= 8) {
  "Dark2"
} else {
  "Set3"  # 'Set3' can handle up to 12 colors
}

# Generate Kaplan-Meier plot with customized options
ggsurvplot(
  fit = km_fit,  # Kaplan-Meier fit
  data = surv_data,  # Data for analysis
  pval = TRUE,  # Display p-value of the log-rank test
  conf.int = TRUE,  # Show confidence intervals
  risk.table = TRUE,  # Include risk table
  risk.table.y.text.col = TRUE,  # Color risk table text
  risk.table.height = 0.25,  # Adjust height of risk table
  xlab = "Time (days)",  # X-axis label
  ylab = "Survival Probability",  # Y-axis label
  break.time.by = 1000,  # Time breaks every 1000 days
  ggtheme = theme_minimal(),  # Minimal plot theme
  palette = palette_choice,  # Use the selected color palette
  legend.title = "AJCC Stage",  # Legend title
  legend.labs = levels(surv_data$ajcc_pathologic_stage),  # Legend labels
  title = "Survival Analysis by AJCC Pathologic Stage"  # Plot title
)

# Perform log-rank test to compare survival distributions by AJCC stages
logrank_test <- survdiff(surv_obj ~ ajcc_pathologic_stage, data = surv_data)

# Display log-rank test results
summary(logrank_test)

```


📉 Key Interpretations:
✅ 1. Clear stage-wise separation:
Survival curves drop at different rates depending on the AJCC stage.

Patients with early-stage cancers (Stage I, IA, IB) have higher survival probabilities over time.

In contrast, Stage IV and III patients have steeper drops, indicating worse survival outcomes.

✅ 2. p-value < 0.0001:
This indicates that the differences in survival between the stages are statistically significant.

In other words, AJCC stage is strongly associated with survival.

✅ 3. Risk Table (below the plot):
Shows the number of patients still under observation ("at risk") at different time intervals for each stage.

Numbers decrease over time as patients either die or are censored (lost to follow-up).

✅ 4. Stage-wise patterns:
Here’s a simplified view of the survival trend:


Stage	Trend
Stage I / IA	Best survival, slowest drop in probability
Stage IB / IIA	Moderate survival
Stage IIB / IIIA / IIIB	Poorer outcomes, quicker drops
Stage IV	Worst prognosis, sharpest decline


```{r}
# Load necessary libraries
library(survival)
library(survminer)
library(ggplot2)

# Convert AJCC stage to character to allow modification
surv_data$ajcc_pathologic_stage <- as.character(surv_data$ajcc_pathologic_stage)

# Merge 'Stage III' into 'Stage IIIA'
surv_data$ajcc_pathologic_stage[surv_data$ajcc_pathologic_stage == "Stage III"] <- "Stage IIIA"

# Convert back to factor and drop unused levels
surv_data$ajcc_pathologic_stage <- droplevels(factor(surv_data$ajcc_pathologic_stage))
surv_data$tobacco_smoking_status <- droplevels(factor(surv_data$tobacco_smoking_status))

# Fit Cox model
cox_fit_merged <- coxph(Surv(survival_time, event) ~ 
                          age_at_diagnosis + 
                          tobacco_smoking_status + 
                          ajcc_pathologic_stage, 
                        data = surv_data)

# Display summary
cox_summary <- summary(cox_fit_merged)

# Create a data frame with HR and confidence intervals
results <- data.frame(
  Variable = rownames(cox_summary$coefficients),
  HR = exp(cox_summary$coefficients[, "coef"]),
  LowerCI = exp(cox_summary$conf.int[, "lower .95"]),
  UpperCI = exp(cox_summary$conf.int[, "upper .95"]),
  PValue = cox_summary$coefficients[, "Pr(>|z|)"]
)

# Forest plot of Hazard Ratios
ggplot(results, aes(x = reorder(Variable, HR), y = HR, ymin = LowerCI, ymax = UpperCI)) +
  geom_pointrange() +
  coord_flip() +
  labs(x = "Variables", y = "Hazard Ratio") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red")

# Kaplan-Meier survival curves (based on the Cox model)
ggsurvplot(survfit(cox_fit_merged), data = surv_data, pval = TRUE)




```

#Random Forest code

```{r}
# Load libraries
library(ranger)
library(survival)
library(ggplot2)

# Fit the random forest survival model
rf_model <- ranger(
  formula = Surv(survival_time, event) ~ age_at_diagnosis + 
                                         tobacco_smoking_status + 
                                         ajcc_pathologic_stage,
  data = surv_data,
  num.trees = 100,           # Start small for testing
  mtry = 2,
  seed = 42,
  importance = "none",
  respect.unordered.factors = "order",
  survival = TRUE,
  keep.inbag = TRUE
)

# Extract survival probabilities for first patient
# These are computed during model training (not in predict)
surv_probs_patient1 <- rf_model$survival[1, ]  # 1st patient
time_points <- rf_model$unique.death.times

# Plot survival curve for Patient 1
df_plot <- data.frame(
  Time = time_points,
  Survival = surv_probs_patient1
)

ggplot(df_plot, aes(x = Time, y = Survival)) +
  geom_line(color = "darkgreen", size = 1.2) +
  labs(
    title = "Predicted Survival Curve (Patient 1)",
    x = "Time",
    y = "Survival Probability"
  ) +
  theme_minimal()
# Number of patients to plot
n_patients <- 5

# Create a long-format data frame for ggplot
surv_plot_df <- data.frame(
  Time = rep(rf_model$unique.death.times, times = n_patients),
  Survival = as.vector(t(rf_model$survival[1:n_patients, ])),
  Patient = factor(rep(1:n_patients, each = length(rf_model$unique.death.times)))
)

ggplot(surv_plot_df, aes(x = Time, y = Survival, color = Patient)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Survival Curves for First 5 Patients",
    x = "Time",
    y = "Survival Probability",
    color = "Patient"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Number of patients to visualize
n_patients <- 5

# Create long-format data frame with survival curves + stage info
surv_plot_df <- data.frame(
  Time = rep(rf_model$unique.death.times, times = n_patients),
  Survival = as.vector(t(rf_model$survival[1:n_patients, ])),
  Patient = factor(rep(1:n_patients, each = length(rf_model$unique.death.times))),
  Stage = rep(surv_data$ajcc_pathologic_stage[1:n_patients], each = length(rf_model$unique.death.times))
)

# Plot survival curves by patient with facet for AJCC stage
library(ggplot2)

ggplot(surv_plot_df, aes(x = Time, y = Survival, color = Patient)) +
  geom_line(size = 1.2) +
  facet_wrap(~ Stage) +
  labs(
    title = "Predicted Survival Curves for First 5 Patients by AJCC Stage",
    x = "Time",
    y = "Survival Probability",
    color = "Patient"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

library(dplyr)
library(ggplot2)
library(tidyr)

# Get survival probabilities from ranger model
surv_probs <- rf_model$survival
time_points <- rf_model$unique.death.times

# Add stage info to each row
stage_vector <- surv_data$ajcc_pathologic_stage
surv_df <- as.data.frame(surv_probs)
surv_df$Stage <- stage_vector

# Group by stage and calculate mean survival at each time point
mean_surv_by_stage <- surv_df %>%
  group_by(Stage) %>%
  summarise(across(.cols = where(is.numeric), .fns = mean, na.rm = TRUE))

# Convert wide to long for plotting
mean_surv_long <- mean_surv_by_stage %>%
  pivot_longer(
    cols = -Stage,
    names_to = "Time_Index",
    values_to = "Survival"
  ) %>%
  mutate(Time = time_points[as.numeric(gsub("V", "", Time_Index))])

# Plot
ggplot(mean_surv_long, aes(x = Time, y = Survival, color = Stage)) +
  geom_line(size = 1.2) +
  labs(
    title = "Average Survival Curves by AJCC Stage",
    x = "Time",
    y = "Survival Probability"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

```{r}
# Load required libraries
library(ranger)
library(iml)
library(dplyr)
library(survival)

# Train the survival random forest model
rf_model <- ranger(Surv(survival_time, event) ~ age_at_diagnosis + tobacco_smoking_status + ajcc_pathologic_stage, 
                   data = surv_data,
                   num.trees = 500,
                   importance = "impurity",
                   mtry = 2,
                   seed = 42)

# Prepare the input data for prediction (excluding survival_time and event)
X <- surv_data %>%
  select(-survival_time, -event)

# Define the prediction function for expected survival time
predict_expected_survival <- function(newdata) {
  pred <- predict(rf_model, data = newdata)
  times <- rf_model$unique.death.times
  
  # Convert to matrix (important when predicting for 1 row)
  survival_matrix <- matrix(pred$survival, nrow = nrow(newdata))
  
  # Compute expected survival time
  expected <- apply(survival_matrix, 1, function(surv) {
    sum(diff(c(0, times)) * surv)
  })
  
  return(expected)
}

# Wrap the function for iml::Predictor compatibility
custom_predict_function <- function(model, newdata) {
  predict_expected_survival(newdata)
}

# Create the Predictor object
predictor <- Predictor$new(
  model = NULL,  # Set to NULL since we use a custom prediction function
  data = X,
  y = surv_data$survival_time,
  predict.fun = custom_predict_function
)

# Inspect input
str(X[1, , drop = FALSE])

# SHAP for the first observation
shap <- Shapley$new(predictor, x.interest = X[1, , drop = FALSE])

# Plot SHAP values
shap$plot()

for (i in 1:5) {  # Just the first 5 patients for example
  shap <- Shapley$new(predictor, x.interest = X[i, , drop = FALSE])
  print(paste("Patient", i))
  print(shap$results)
  plot(shap)
}

View(surv_data)

```